Digit Recognition: Nearest Neighbors vs. Perceptron

### 2013-06-11 Paper/paper.tex
paper.tex

The first method we implemented to classify our testing sample was a simple
majority voting algorithm which selects the majority label from the \textit{K}
closest neighbors. For example, if we ran the nearest neighbors algorithm with
$K = 5$ and 3 of the neighbors had a label of 6, we would classify our testing
sample as a 6.

@@

The weight function is an inverse distance function of the form:

@@

@@

The perceptron algorithm attempts to separate a training data set into two
classes and attempts to classify future testing samples as belonging to one of
these two classes. It does this by storing a list of all the mistakes it has
made previously and using this knowledge to adjust future predictions in the
hopes of improving classification accuracy.

@@

One problem that needs to be addressed is that a standard perceptron is a binary
classifier meaning it selects between one of two classes. We resolved this issue
by training $10 \choose 2$ $=45$ classifiers. Each classifier is responsible for
classifying between two digits. For example, we would train a perceptron to
classify between 0 and 1, and a second to classify between 0 and 2, and so
on. When we classify we vote between all of our trained classifiers and output
the majority element.

@@

Both of these kernels have a parameter which needs to be tuned, $d$ in the case of
the polynomial kernel and $\sigma$ in the case of the exponential kernel. We
used cross validation to find out which kernel gave us the best result as well
as which value of $d$ and $\sigma$ was optimal.

\begin{table}[ht]
\textbf{\caption{Perceptron Cross Validation Success Rates}} % title
\centering
\begin{tabular}{c | c c }
Kernel & $d, \sigma = 5$ & $d, \sigma = 10$ \\
\hline
Polynomial & 95.67\% & 94.52\% \\
Exponential & 93.84\% & 96.3\% \\ [1ex]
\end{tabular}
\label{table:perceptron}
\end{table}

\begin{center}
\includegraphics[width=406px]{Perceptron_CrossValidation.png}
\end{center}

@@

Our cross validation indicates that the exponential kernel with a $\sigma$ of 10
leads to the highest success rate of 96.3\%. An interesting observation is that
our polynomial kernel actually became less successful when it's dimension was
increased whereas the exponential kernel got better. We suspect that this is
because the polynomial kernel began to overfit the training data.

Finally, the runtime of the perceptron algorithm is a little over an hour. It
takes approximately 45 minutes to train and about 20 minutes to test the
data. The faster runtime allowed us to cross validate on the full data set so
our cross validation results are more accurate than those we obtained for KNN.

@@

\begin{table}[ht]
\textbf{\caption{KNN vs. Perceptron on Full Testing Set}}
\centering
\begin{tabular}{ c| c c}
  \hline
  Algorithm & Percent Success & runtime\\
  \hline
  Nearest Neighbors, $k = 4$   & 96.857\% & $\sim$8 hours\\
  Perceptron   & 96.300\%  & $\sim$1 hour\\
\end{tabular}
\end{table}

Table 4 summarizes the accuracy and runtimes of weighted KNN with $K = 4$ and
perceptron with the polynomial kernel and $d = 10$. KNN is clearly more
accurate, however because perceptron is nearly 8x faster, there could be many
cases where it is a better choice.

@@

One reason why KNN may have been so successful in this trial is due to the large
training data set and small number of classes to classify. With only 10 posible
classifications, there are likely around 4,2000 training samples of each
number. The odds of at least a few of those being similar to the test image
intuitively seem high.

An explanation for the perceptron algorithm being less successful than KNN is
that it may have overfit our training data because there was no
regularization. This could explain why our polynomial kernel did worse with a
higher dimension kernel function.

>>>>

The first method we implemented to classify our testing sample was a simple
majority voting algorithm which selects the majority label from the \textit{K}
closest neighbors. For example, if we ran the nearest neighbors algorithm with
$K = 5$ and 3 of the neighbors had a label of 6, we would classify our testing
sample as a 6. If there is a tie between two labels then the label is picked 
arbitrarily.

@@

Many different weight functions are available but we chose to use the inverse
distance. This allows samples that are closest to one another to have the 
largest impact while discounting those that are farthest away. The distance
formula we used is below:

@@

\subsection{Analyzation}

The KNN algorithm is often very accurate but can be too slow for some applications. 
In the case of OCR it proved to be the easiest to implement and the most accurate. 
The algorithm uses a brute force tactic to compare a point against all other training
points.  This allows a decision to be made with all the information available.  However,
we were still only able to obtain a max of 96.857\% accuracy.  There are several possible
reasons for this.  First, OCR is very difficult.  There are likely some hand written samples
that simply do not have a good representation of nearest neighbors. Peoples handwriting 
styles are different enough that some of the more unique interpretations of numbers did 
not get classified. Second, we only ran cross validation on a small subset of the samples.  
We did not have the time to run it on the full 42,000 samples and thus the k value we used 
for testing may not have been optimal.

@@

The perceptron algorithm is a binary classifier that minimizes the hinge loss objective
function.  Typically, this would require storing weights to use in making a prediction 
during testing. However, the kerneled version of the perceptron instead stores the
mistakes made during training to use during testing. All the mistakes made during
training are then passed to a kernel with the testing data and test time in order to
make a prediction of the testing label.

@@

One problem that needed to be addressed is that a standard perceptron is a binary
classifier meaning it selects between one of two classes. We resolved this issue
by training $10 \choose 2$ $=45$ classifiers. Each classifier is responsible for
classifying between two digits. For example, we would train a perceptron to
classify between 0 and 1, and a second to classify between 0 and 2, and so
on. When we classify we vote between all of our trained classifiers and output
the majority element.

@@

Both of these kernels have a parameter which needs to be tuned, $d$ in the case of
the polynomial kernel and $\sigma$ in the case of the exponential kernel. We
used 10-fold cross validation to find out which kernel gave us the best result as well
as which value of $d$ and $\sigma$ was optimal. Below are the results of the cross
validation:

\begin{center}
\includegraphics[scale=.7]{cv.png}
\end{center}

There are many different possible selections of kernels to use and our implementation
allows for the user to pass a command line argument to select the kernel and dimension.
We chose these two because we knew the performance was adequate based on our
previous homework. Due to time constraints we were not able to cross validate and expire
any other kernels.  We briefly explore the Gaussian but the mistake files grew to over 
1Gb in size and thus the training slowed considerably.

@@

Our cross validation indicates that the exponential kernel with a $\sigma$ of 15
leads to the highest success rate of 96.3\%. An interesting observation is that
our polynomial kernel actually became less successful when it's dimension was
increased whereas the exponential kernel got better. We suspect that this is
because the polynomial kernel began to overfit the training data.

Finally, the runtime of the perceptron algorithm is a little over an hour. We timed
a particular train and test run with the exponential kernel and found that it trained
in 21 minutes and tested in 48 minutes.  This was similar to what we observers
with the polynomial kernel as well. The faster runtime allowed us to cross validate 
on the full data set so our cross validation results are more accurate than those 
we obtained for KNN.

@@

\begin{table}[ht]
\textbf{\caption{KNN vs. Perceptron on Full Testing Set}}
\centering
\begin{tabular}{ c| c c}
  \hline
  Algorithm & Percent Success & runtime\\
  \hline
  Nearest Neighbors, $k = 4$   & 96.857\% & $\sim$8 hours\\
  Perceptron   & 96.129\%  & $\sim$1 hour\\
\end{tabular}
\end{table}

Table 4 summarizes the testing accuracy and runtimes of weighted KNN with $K = 4$ and
perceptron with the exponential kernel and $\sigma = 15$. KNN is clearly more
accurate, however because perceptron is nearly 8x faster, there could be many
cases where it is a better choice.

@@

One reason why KNN may have been so successful in this trial is due to the large
training data set and small number of classes to classify. With only 10 possible
classifications, and 42,000 training samples, the odds of at least a few of those
being very similar to the test image intuitively seem high. As briefly mentioned earlier
there will be a small minority of strangely written numbers that will not have a good
match to any of the training samples and thus will not be classified properly.

An explanation for the perceptron algorithm being less successful than KNN is
that it may have overfit our training data because there was no
regularization. This could explain why our polynomial kernel did worse with a
higher dimension kernel function. We would have liked to explore a kerneled
SVM to resolve this issue but we ran into problems with derivation and implementation.
Due to time constraints we stuck with the perceptron and explored its implementation
further through cross validation and kernel implemenations.
